Fast and Flexible Binary Instrumentation for the Kernel

### 2013-11-27 VEE-14/paper.tex
Did some stuff.. which I don't remember
Described indirect branch lookup, resectioned some stuff.

\paragraph{Function return addresses}\label{para:return_address_transparency} Unlike most DBT frameworks, Granary inlines \texttt{call} instructions into basic blocks, which exposes code cache addresses to instrumented module code in the form of return addresses. Inlining \texttt{call} instructions avoids return-address mispredictions and unnecessary control-flow by building longer basic blocks.

\paragraph{Interrupt return addresses} Kernel interrupts are not sensitive to return addresses and hence relaxing return address transparency has no effect on interrupts. A special case arises for page fault exceptions that occur within specific kernel functions that access user space data \cite{btkernel}. The Linux kernel records ranges of code addresses that are permitted to fault in an ``exception table'' data structure, and on a page fault exception, checks if the interrupt return address belongs to one of the pre-defined ranges. Granary records any exception table entries as part of each basic block's meta-data, so that a page fault in a basic block can be mapped to the correct native faulting program counter.

@@

\subsection{Reentrancy}

\paragraph{Code cache}
Granary's code cache and JIT translation mechanism is fully reentrant. Context-switches into Granary disable interrupts and switch execution onto a CPU-private stack. Context switches out of Granary and into the code cache or native code returns execution to the native stack and re-enables interrupts. This approach presents a minor challenge: context-switches save or restore the entire register state, so as not to disturb the execution of the instrumented program. Therefore, if a context switch into Granary occurs because the next basic block must be found/translated, then the address of that next basic block must be stored \emph{somewhere} in order for execution to eventually reach that basic block when the context switches back to the code cache. But context switching is not allowed to perturb the register state of the instrumented program, and so the target address must be resolved by another means.

Like btkernel, Granary solves this problem of figuring out how to get to where the code needs to go  by generating ``edge'' code for each direct control-flow instruction (e.g. \texttt{call}, \texttt{jmp}, \texttt{jcc}). When a basic block is first translated, its direct control-flow instructions (CTIs) are converted into jumps to CTI-specific edge code. Edge code lives outside the code cache and is responsible for transferring control from a basic block ($B_{pred}$) to Granary so that the next basic block ($B_{succ}$) can be found/translated, and then connected ($B_{pred} \underset{\texttt{jz}}{\to} B_{succ}$) via hot code patching. Reentrancy of edge code is obtained by spilling registers onto the runtime call stack, and later restoring them before jumping back to the now-patched CTI (\Cref{fig:direct_edge_code}).

Granary also generates edge code for indirect control-flow instructions; however, this edge code is split into two separate components: entry and exit edge code. Entry edge code is generated on a per-indirect CTI instruction basis, and spills registers onto the runtime call stack. Entry edge is responsible for indirect branch lookup, and will indirectly jump to the exit edge code for a given basic block. Exit edge code is generated for each basic block that is targeted by an indirect CTI. Exit edge code restores the register state saved by the entry edge code, and then directly jumps to the beginning of the targeted basic block.

\paragraph{Instrumentation}

% Code block showing an example JZ (jump if ZF=1).
\lstset{language=[x64]Assembler}
\newsavebox\nativejcc
\begin{lrbox}{\nativejcc}
\begin{minipage}[b]{4cm}
\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
cmp $0x0, %rax
jz value_is_zero 
...
( fall through )
\end{lstlisting}
\end{minipage}
\end{lrbox}

\begin{figure}[t!]
\subfloat[Native Code]{\usebox\nativejcc}
\hfill
 \subfloat[Translated Code]{\usebox\translatedjcc}
\caption{\label{fig:direct_edge_code}Example translation of a conditional control flow instruction (\texttt{jz}) and its edge code. The original \texttt{jz} is translated into a \texttt{jz} that uses a 32-bit relative displacement to transfer control to an edge code function call (\texttt{edge\_jz}). When executed, the function call reads its target from a data structure (\texttt{patch\_edge\_jz}), which contains all information necessary to resolve the \texttt{jz}'s target and patch the original \texttt{jz}. Patching is performed with a four byte atomic write to the relative displacement operand of the \texttt{jz} instruction. Finally, the return address of the \texttt{call} instruction is altered to transfer control to the newly resolved target.}
\end{figure}

@@

\subsection{Optimizations}
\paragraph{Ahead-of-time Translation} \label{sec:aot}
Granary ends basic blocks at conditional branches (\texttt{jCC}), direct and indirect \texttt{jmp}s, and \texttt{ret}s and \texttt{iret}s. Basic blocks ending in conditional branches are extended to contain two trailing CTIs: a copy of the native \texttt{jCC}, and a synthesized direct \texttt{jmp} to the code executed when the condition code is not set (i.e. when the conditional jump is not taken).

@@

\paragraph{Patching and Alignment}

@@

\paragraph{Trace Allocation}
Arranging for commonly executed sequences of basic blocks to appear adjacent in memory, also called tracing, typically improves the runtime performance of instrumented programs \cite{DynamoRIO}. Typical approaches to tracing require a runtime profiling infrastructure (e.g. recording basic block execution counts). Granary attempts to reap the benefits of tracing without the runtime profiling overhead by opportunistically forming traces using ahead-of-time translation (\Cref{sec:aot}), and by arranging for related basic blocks to be allocated sequentially in memory.

@@

%, and is associated with a private memory pool from which to allocate memory for basic blocks. The net result is that most of the basic blocks eecuted in service of a single system call are allocated from that system call's memory pool.

%The intuition for this approach is that kernel code is not very ``loopy", and so

%Granary services memory allocation requests for basic block instructions from memory pools. By default, Granary has as many memory pools as there are CPUs. However, when trace allocation is enabled

%Tracing is also beneficial for

%However, tracing of this sort requires mechnisms for evicting stale basic blocks,

%Constructing traces of ``hot'' 
%JIT generally results in better performance .

%By default, Granary allocates the memory needed to hold the instructions of a basic block from CPU-private memory pools. Each memory pool serves basic block allocation requests out of fixed-size memory blocks, ranging from 4K to 24K in size.

\paragraph{Indirect Control Flow}

@@

\paragraph{Probe-based instrumentation}

@@

\paragraph{Wrapper-based instrumentation}

>>>>

\subsubsection{Function return addresses}\label{para:return_address_transparency} Unlike most DBT frameworks, Granary inlines \texttt{call} instructions into basic blocks, which exposes code cache addresses to instrumented module code in the form of return addresses. Inlining \texttt{call} instructions avoids return-address mispredictions and unnecessary control-flow by building longer basic blocks.

\subsubsection{Interrupt return addresses} Kernel interrupts are not sensitive to return addresses and hence relaxing return address transparency has no effect on interrupts. A special case arises for page fault exceptions that occur within specific kernel functions that access user space data \cite{btkernel}. The Linux kernel records ranges of code addresses that are permitted to fault in an ``exception table'' data structure, and on a page fault exception, checks if the interrupt return address belongs to one of the pre-defined ranges. Granary records any exception table entries as part of each basic block's meta-data, so that a page fault in a basic block can be mapped to the correct native faulting program counter.

@@

\subsection{Reentrancy}
Reentrancy is an important concern for all DBT systems, and is especially important for DBT systems operating in the kernel environment. For example, hardware interrupts can redirect control flow at any instruction, and interrupt delivery cannot be delayed past the execution of memory instructions that might alter the interrupt delivery state \cite{DRK}. Preserving interrupt transparency, however, requires that interrupts be delayed until the end of any injected instrumentation instructions.  Together, these requirements make it very difficult to write instrumentation tools.

\subsubsection{Code cache}
Granary's code cache and JIT translation mechanisms are fully reentrant. Context-switches into Granary disable interrupts and switch execution onto a CPU-private stack. Context switches out of Granary and into the code cache or native code returns execution to the native stack and re-enables interrupts. This approach presents a minor challenge: context-switches save or restore the entire register state, so as not to disturb the execution of the instrumented program.

%Therefore, if a context switch into Granary occurs because the next basic block must be found/translated, then the address of that next basic block must be stored \emph{somewhere} in order for execution to eventually reach that basic block when the context switches back to the code cache. But context switching is not allowed to perturb the register state of the instrumented program, and so the target address must be resolved by another means.

%Like btkernel, Granary solves this problem of figuring out how to get to where the code needs to go  by generating ``edge'' code for each direct control-flow instruction (e.g. \texttt{call}, \texttt{jmp}, \texttt{jcc}).

%Granary also generates edge code for indirect control-flow instructions; however, this edge code is split into two separate components: entry and exit edge code. Entry edge code is generated on a per-indirect CTI instruction basis, and spills registers onto the runtime call stack. Entry edge is responsible for indirect branch lookup, and will indirectly jump to the exit edge code for a given basic block. Exit edge code is generated for each basic block that is targeted by an indirect CTI. Exit edge code restores the register state saved by the entry edge code, and then directly jumps to the beginning of the targeted basic block.

\subsubsection{Instrumentation}

@@

\subsection{Implementation}
\subsubsection{Direct Control Flow}
% Code block showing an example JZ (jump if ZF=1).
\lstset{language=[x64]Assembler}
\newsavebox\nativejcc
\begin{lrbox}{\nativejcc}
\begin{minipage}[b]{4cm}
\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
cmp $0x0, %rax
jz value_is_zero 
...
( fall through )
\end{lstlisting}
\end{minipage}
\end{lrbox}

\begin{figure}[t!]
\subfloat[Native Code]{\usebox\nativejcc}
\hfill
 \subfloat[Translated Code]{\usebox\translatedjcc}
\caption{\label{fig:direct_edge_code}Example translation of a conditional control flow instruction (\texttt{jz}) and its edge code. The original \texttt{jz} is translated into a \texttt{jz} that uses a 32-bit relative displacement to transfer control to an edge code function call (\texttt{edge\_jz}). When executed, the function call reads its target from a data structure (\texttt{patch\_edge\_jz}), which contains all information necessary to resolve the \texttt{jz}'s target and patch the original \texttt{jz}. Patching is performed with a four byte atomic write to the relative displacement operand of the \texttt{jz} instruction. Finally, the return address of the \texttt{call} instruction is altered to transfer control to the newly resolved target.}
\end{figure}
Like btkernel, Granary translates direct control-flow instructions (CTIs) into equivalent jumps into CTI-specific ``edge code" (\Cref{fig:direct_edge_code}). Edge code lives outside the code cache, so as to not pollute the icache \cite{btkernel}.

When control flows through a basic block $B_{pred}$ for the first time, any jumps taken will transfer control to edge code that is specific to the jump taken. Edge code contains only a single indirect \texttt{call} instruction, thus allowing multiple edge codes to occupy a single cache line. The indirect \texttt{call} transfers control to Granary via a function pointer located in an ``edge data structure", which describes how to translate the target basic block, $B_{succ}$. Granary gains control and looks up or translates the target code according instrumentation policy information stored in the edge data structure. Then, Granary uses hot-code patching (\Cref{sec:patching}) to modify the original CTI that led to the edge code \texttt{call}, so that it now transfers control to $B_{succ}$ instead. Finally, Granary releases control and returns to $B_{succ}$.

% and is responsible for transferring control from a basic block ($B_{pred}$) to Granary so that the next basic block ($B_{succ}$) can be found/translated, and then connected ($B_{pred} \underset{\texttt{jz}}{\to} B_{succ}$) via hot code patching.

\subsubsection{Indirect Control Flow}

%Indirect branch lookup through a global, fixed-size hash table. A basic block making an indirect control flow transfer jumps to an in-edge that knows where to find the indirect CTI target (e.g. in a specific register or memory location). The in-edge then uses the target as an index into the hash table, and jumps to the target address stored in the indexed hash table entry. That entry will either be the address of an out-edge, or of Granary's indirect branch resolver. Each out-edge is specific to a target basic block, and first checks if the intended jump target matches the out-edge's target. If so, then control is transfered to the target basic block, otherwise control is transfered either to Granary or to the next out-edge. Granary chains together out-edges through conditional jumps just as

%Hash table entries are code addresses of out-edge code stub. Each out-edge code stub is specific to a target basic block, and is responsible for checking if the intended control-flow target is the out-edge's target or not. If not, the out-edge either directly jumps to the next out-edge whose target also fits into the same hash table bucket, or jumps to Granary to determine the target basic block.

Granary translates indirect control flow instructions (\texttt{call}, \texttt{jmp}) into their equivalent direct control flow versions, where the direct versions jump to ``in-edge" code. In-edge code loads the indirect jump target (e.g. a specific register or memory location), saves the flags register, and several scratch registers. Finally, the in-edge uses the indirect target as an index into a global, fixed-size hash table of indirect branch targets. The entries of the hash table either point to Granary or to ``out-edge" code (\Cref{fig:ibl}).

\begin{figure}[t!]
\hspace{4em}\includegraphics[width=0.9\columnwidth]{diagrams/ibl.pdf}\vspace{-3em}
\caption{\label{fig:ibl}Process of resolving an indirect branch. Control transfers from a source basic block, to the in-edge code specific to where the indirect target is stored (e.g. register, memory address), through the global hash table and on to either one or more out-edge code stubs, or to Granary.}
\end{figure}

Out-edge code is specific to a target basic block, and performs the function of separate chaining in the indirect branch lookup hash table. When control transfers to out-edge code, the out-edge checks to see if the indirect target matches the out-edge code's target basic block. If they match, then the out-edge code restores scratch and flags registers and jumps directly to the target basic block. Otherwise, the out-edge code jumps to another out-edge code or to Granary. If one out-edge code block jumps to another one, then the targets of the two out-edge code blocks collide in the global hash table. In this respect, out-edge code is chained together through control flow, just as colliding entries in a data hash table can be chained together through a linked list.

%If an entry points to Granary, then no as-of-yet resolved indirect jump target hashes to that entry.

%Granary implements separate chaining through code (i.e. control flow jumps, as opposed to date, e.g. linked list pointers)
%In the event that one-or-more indirect jump targets do hash to an existing entry
%In-edge code is responsible for 
% to first search for their targets in a global, fixed-size hash table, and failing that, context-switch into Granary to perform a full lookup and potential translation.
%using a fixed-size hash table, where the native target of the indirect

\subsection{Optimizations}
\subsubsection{Ahead-of-time Translation} \label{sec:aot}
Granary ends basic blocks at conditional branches (\texttt{jCC}), direct and indirect \texttt{jmp}s, and \texttt{ret}s and \texttt{iret}s. Basic blocks ending in conditional branches are extended to contain two trailing CTIs: a copy of the native \texttt{jCC}, and a synthesized direct \texttt{jmp} to the code executed when the condition code is not set (i.e. when the conditional jump is not taken).

@@

\subsubsection{Patching and Alignment}\label{sec:patching}

@@

\subsubsection{Trace Allocation}
Arranging for commonly executed sequences of basic blocks to appear adjacent in memory, also called tracing, typically improves the runtime performance of instrumented programs \cite{DynamoRIO}. Typical approaches to tracing require a runtime profiling infrastructure (e.g. recording basic block execution counts). Granary attempts to reap the benefits of tracing without the runtime profiling overhead by opportunistically forming traces using ahead-of-time translation (\Cref{sec:aot}), and by arranging for related basic blocks to be allocated sequentially in memory.

@@

%, and is associated with a private memory pool from which to allocate memory for basic blocks. The net result is that most of the basic blocks eecuted in service of a single system call are allocated from that system call's memory pool.
%The intuition for this approach is that kernel code is not very ``loopy", and so 
%Granary services memory allocation requests for basic block instructions from memory pools. By default, Granary has as many memory pools as there are CPUs. However, when trace allocation is enabled
%Tracing is also beneficial for 
%However, tracing of this sort requires mechnisms for evicting stale basic blocks, 
%Constructing traces of ``hot'' 
%JIT generally results in better performance .
%By default, Granary allocates the memory needed to hold the instructions of a basic block from CPU-private memory pools. Each memory pool serves basic block allocation requests out of fixed-size memory blocks, ranging from 4K to 24K in size.

\subsubsection{System Calls}

@@

\subsubsection{Probe-based instrumentation}

@@

\subsubsection{Wrapper-based instrumentation}
